---
title: "EDA of Obesity Data"
author: "Rafael Pilliard Hellwig"
date: "12/01/2021"
output: 
  html_document:
    self_contained: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r load-pkg, warning=FALSE, message=FALSE}
# Load packages
library(tidyverse)
library(WDI)

# silence dplyr summarise warnings
options(dplyr.summarise.inform = FALSE)
```


Let's start by reading-in the obesity data (using the pre-cleaned `data/obesity-cleaned.csv` file).

```{r read-obesity-cleaned}
# Read-in and tidy obesity data
file <- here::here("data", "obesity-cleaned.csv")
colspec <- cols(
  X1 = col_skip(),
  Country = col_character(),
  Year = col_double(),
  `Obesity (%)` = col_character(),
  Sex = col_character()
)
obesity <- read_csv(file, col_types = colspec) %>%
  janitor::clean_names() %>%
  filter(sex != "Both sexes") %>%
  filter(obesity_percent != "No data") %>%
  mutate(rate_obesity = as.numeric(str_extract(obesity_percent, "^\\d+\\.\\d")) / 100) %>%
  select(-obesity_percent)
```


Next, let's download some data from the World Bank. We'll use (sex-disaggregated) indicators on percentage completion of basic education, smoking rates, and population.

```{r get-wb-data, cache=TRUE}
# List World Bank indicators
indicators <- c(rate_pcr.Male     = 'SE.PRM.CMPT.MA.ZS',
                rate_pcr.Female   = 'SE.PRM.CMPT.FE.ZS',
                rate_smoke.Male   = "SH.PRV.SMOK.MA",
                rate_smoke.Female = "SH.PRV.SMOK.FE",
                pop.Male          = 'SP.POP.TOTL.MA.IN',
                pop.Female        = 'SP.POP.TOTL.FE.IN')

# Download World Bank data
wb <- WDI(indicator = indicators, extra = TRUE)
```

Next, we combine both datasets and ensure that all variables are converted to counts. This ensures that each row is a proper stratum that can be aggregated any way.


```{r reshape-wb}
# Reshape world bank values to counts
wb_df <- wb %>%
  as_tibble %>%
  pivot_longer(contains("."), names_to = c("indicator", "sex"),
               names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = indicator, values_from =  value) %>%
  pivot_longer(starts_with("rate_"), 
               names_to = "indicator",
               values_to = "rate",
               names_pattern = "_(.*)$") %>%
  drop_na(rate) %>%
  mutate(across(rate, ~ . / 100),
         count = round(pop * rate)) %>%
  select(-rate) %>%
  pivot_wider(names_from = indicator, values_from = count)

# Combine data
combo <- obesity %>%
  left_join(wb_df, by = c("country", "year", "sex")) %>%
  mutate(obese = rate_obesity * pop) %>%
  drop_na(pop) %>%
  select(-rate_obesity)
```


Finally, for convenience, let's create a dataset containing the latest data:

```{r create-combo-latest}
# create a version that only contains the latest year
combo_latest <- combo %>%
  arrange(desc(year)) %>%
  group_by(country, sex) %>%
  slice(1) %>%
  ungroup
```

Let's try visualizing a few of the relationships. Let's start with looking for an association between smoking and obesity:


```{r plot-obesity-smoking, warning=FALSE, message=FALSE}
combo_latest %>%
  group_by(region, country) %>%
  summarise(smoke_rate = sum(smoke) / sum(pop),
            obesity_rate = sum(obese) / sum(pop)) %>%
  ggplot(aes(x = smoke_rate, y = obesity_rate)) +
  geom_point(aes(colour = region)) +
  geom_smooth(method = "lm") +
  labs(title = "Positive Relationship Between Smoking and Obesity",
       x = "Rate of Smoking",
       y = "Rate of Obesity",
       colour = "Region")
```

Next, let's look at how obesity has changed over time.

```{r plot-obesity-ts, warning=FALSE, message=FALSE}
combo %>%
  group_by(region, year) %>%
  summarise(obesity_rate = sum(obese, na.rm = TRUE) / sum(pop, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = obesity_rate, colour = region)) +
  geom_line() +
  labs(title = "Obesity Rates Over Time, by Region",
       x = "Year",
       y = "Rate of Obesity",
       colour = "Region")
```

A few of the values are a bit erratic. That's something to look into later.

